---
title: "Can Colleges Afford to Pay Their Players?"
author: "Dylan Hunt"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r}
# load packages and read in your data. 
library(tidyverse)
library(mdsr)
library(dplyr)
library(ggplot2)
NCAA_Financials <- read_csv("pubDataRepo.csv")
#head(NCAA_Financials)  #Remove the first # to display this for your proposal. Don't display in your final report. 
```

  Do colleges have the money to pay their players? This question has become very relevant in collegiate sports today. Some universities are bringing in almost 100 million dollars in revenue these days. Some argue that because players are bringing in this money, they should get paid more than their scholarships. Currently, athletes can sign name image and likeness (NLI) deals, but colleges cannot directly pay athletes. So, athletes can sell themselves for marketing to companies or individuals, but they are not seeing any of the revenue colleges are bringing in. The Huffington Post (https://projects.huffingtonpost.com/ncaa/reporters-note) collected data on the economics of college athletics. Their data was from 28 division one conferences from 2010 to 2014. This article will discuss how this data from The Huffington Post gives a glimpse into an answer to this question.

```{r, fig.height=8, fig.width=10}
NCAA <- NCAA_Financials %>% group_by(conference) %>% summarize(avgsub=mean(inflation_adjusted_subsidy),avgendow=mean(inflation_adjusted_endowments), avgathexp=mean(inflation_adjusted_athletic_expenses),avgticket=mean(ticket_sales),avgathrev=mean(inflation_adjusted_athletic_revenues), avgnetrev=mean(inflation_adjusted_net_revenue), avgenroll=mean(full_time_enrollment))

NCAA <- arrange(NCAA,desc(avgathrev),by_group = FALSE)

confnum <- 1:28

NCAA <- cbind(NCAA,confnum)

NCAA$conference <- factor(NCAA$conference, levels = c("Southeastern Conference","Big Ten Conference", "Big 12 Conference","Atlantic Coast Conference","Pacific-12 Conference","American Athletic Conference","Mountain West Conference","Mid-American Conference","Conference USA","Atlantic 10 Conference","Big West Conference","Colonial Athletic Association","America East Conference","Missouri Valley Conference","Sun Belt Conference","Western Athletic Conference","Big Sky Conference","Northeast Conference","Southern Conference","Independent","The Summit League","Horizon League","Ohio Valley Conference","Southland Conference","Atlantic Sun Conference","Big South Conference","Mid-Eastern Athletic Conference","Southwestern Athletic Conference"), labels= c("SEC","Big 10","Big 12","ACC","Pac-12","American","Mountain West","Mid-American","Conference USA","Atlantic 10","Big West","Colonial","America East","Missouri Valley","Sun Belt","Western","Big Sky","Northeast","Southern","Independent","Summit League","Horizon","Ohio Valley","Southland","Atlantic Sun","Big South","Mid-Eastern","Southwestern"))

NCAA <- mutate(NCAA,Power_Five=ifelse((conference =="SEC"|conference=="Big 10"|conference=="Big 12"|conference=="ACC"|conference=="Pac-12"),T,F))

ggplot(data=NCAA, aes(x=conference)) +
  geom_bar(mapping=aes(x=fct_reorder(.f=conference, .x=avgathrev/1000000, .desc=TRUE), y=avgathrev/1000000, fill=Power_Five), stat = "identity") + xlab("Conference") + ylab("Millions of Dollars") + ggtitle("Different Conferences Average Revenue vs. Average Expenses") + theme(axis.text.x = element_text(angle = 90),legend.position = "none") + scale_fill_manual(values = c('#00FF00', '#009900')) + geom_line(mapping=aes(y=avgathexp/1000000,x=confnum),color="red", size=2) +
geom_point(mapping=aes(y=avgathexp/1000000,x=confnum),color="red") +
geom_curve(x="Atlantic 10", xend="American", y=75, yend=50, arrow=arrow(length=unit(0.3, "cm")), curvature=0.5, color="#993300") + geom_text(x="America East",y=75,label="Average Expences", color="#993300") + geom_curve(x="Pac-12", xend="Big 12", y=93, yend=95, arrow=arrow(length=unit(0.3, "cm")), curvature=0.5,color="#009900") + geom_text(x="Mountain West",y=90, label="Power 5 Conferences", color="#009900") +
geom_curve(x="Big Sky", xend="Missouri Valley", y=40, yend=5, arrow=arrow(length=unit(0.3, "cm")), curvature=0.5,color="#003300") + geom_text(x="Independent",y=40, label="Average Revenue", color="#003300")
```

  Many variables come into play when answering the question if schools can afford to pay their players. Do schools’ athletic expenses equal their revenue? Are they making a profit? Are they losing money from their athletics? Athletic departments, especially at the division one level, have many expenses. They have the pay for coaches, trainers, facilities, travel, hotels, etc. But they also have a few ways of obtaining revenue. Ticket sales, merchandise sales, and tv revenue are just a few ways that athletic departments produce their revenue. This relationship between revenue and expense is how we will examine if athletic departments have the monetary ability to pay their players.

  We investigated both athletic revenue and expenses from these 5 years. Both variables were inflation adjusted to give an accurate representation of these variables over the years. Additionally, all universities were filtered into athletic conference. This step occurred because each conference is different and has different circumstances. For instance, some conferences like the Big 10 have their own tv network making their revenue system different from others. Five conferences known as the power five were highlighted also. These conferences are the best conferences in competition and are the most widely known. It would be unfair to combine all schools, due to the different circumstances that exist among schools.

  After filtering out by conference we see a discrepancy in revenue generated by different conferences. Also, expenses were seen to decrease as revenue decreased, showing a direct relationship. Some conferences expenses were seen as essentially equal to their revenue produced. Some schools on average were not making a distinguishable profit. But others, like the power five, were seen to have made some profit. This is visible in the area above the expenses line. These findings tell us that some conferences do have the requisite revenue to pay its players. But some conferences do not have the money to spare. This study produces a mixed result and gives an example as to why this issue is a complex issue. Some universities can afford to pay their players, but other universities do not have the money to pay them.
